---
title: "broadcasting"
output: html_document
---

```{r}
base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
```

In this tutorial we explore a basic system for broadcasting, which is useful
for quickly constructing ensemble runs with different combinations of model 
inputs. This functionality is built into the `EnsembleInputBroadcast` class, 
though this walk through focuses only on the underlying broadcast functions.


# Illustrative Examples

To start, consider an example where we want to run a different initial condition
ensemble at a number of sites. Each site has different initial conditions, and 
we consider running two ensemble members per site. The total number of model 
runs in this case is thus six (three sites times two ensemble members per site).
The two dimensions ("slots") in this example are `site` and `ic`. The below
code first defines the unique set of values for each slot, consisting of 3
sites and 6 initial conditions. The function `rule_recycle()` is used to 
repeat the sequence of three sites twice to fill out a complete `(6,2)` matrix
of slot combinations.

```{r}
slots <- list(
  site = list("s1", "s2", "s3"),
  ic = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2")
)

lens <- vapply(slots, length, integer(1))
idcs <- rule_recycle(lens)
idx_names <- visualize_slot_grid(idcs, names(slots))

print(idcs)
print(idx_names)
```
We have also used the function `visualize_slot_grid()` to fill in the indices
with names to make it easier to read. Note that the names are the slot names
with indices appended, not the slot values themselves.

Let's now extend this example. Suppose that we want to extend the runs from 
above so that each run is repeated for multiple (site-independent) parameter
values. If we consider a parameter ensemble of size two then there will be 
twelve runs in total (3 sites times 6 initial conditions times 2 parameter sets).
In this case we can use the `rule_recycle()` function as before on the first 
two dimensions, then take the cartesian product of the result with the 
parameter ensemble. We can construct a new broadcast rule that performs both
of these steps using the `get_composite_rule()` function.

```{r}
slots <- list(
  site = list("s1", "s2", "s3"),
  ic = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)

# Create new broadcast rule
new_rule <- get_composite_rule(groups=list(c(1,2), c(3)), # slot groups: (site, ic), (param)
                               rules=list(rule_recycle, rule_match))

# Apply rule to slot values
lens <- vapply(slots, length, integer(1))
idcs <- new_rule(lens)
idx_names <- visualize_slot_grid(idcs, names(slots))

print(idcs)
print(idx_names)
```

Observe that this repeats all of the same combinations from the first example 
once per each parameter set.

# Broadcast Rules

As illustrated above, the broadcast rule functions simply take in integer vectors
corresponding to array dimensions and return a matrix of shape 
`(n_combinations, n_dims)`. We walk through rule functions that are already
implemented. For special cases, ad hoc rule functions can easily be constructed.

## Cartesian product
Constructs every possible combination.

```{r}
rule_cartesian(c(2,3,1))
```

## Parallel Match
This simply lines up the dimensions; the first combination consists of the
first value in each dimension, and so on. This function requires the lengths
of every dimension to be equal.

```{r}
idx <- rule_match(c(2,2,2))
print(idx)

# Error, non-equal dimension lengths
idx <- try(rule_match(c(2,3,1)))
idx
```
## Broadcasting
`rule_braodcast()` actually implements the standard notion of broadcasting, as
used in packages such as NumPy or rray. Each dimension length in this case must
be 1 or equal to the maximal dimension length. The unit dimensions are repeated
to equal the maximal length.

```{r}
idx <- rule_broadcast(c(1,2))
print(idx)

idx <- rule_broadcast(c(3,1,3))
print(idx)

# Invalid
idx <- try(rule_broadcast(c(3,1,2)))
idx
```

## Recycling
The recycling rule simply repeats the pattern in a shorter dimension until it 
lines up with the longer dimension. This is similar to the way R cycles values
but is safer, since it will throw an error if the shorter dimensions don't 
evenly divide the maximal dimension length.

```{r}
idx <- rule_recycle(c(2, 4))
print(idx)

idx <- rule_recycle(c(2, 4, 8))
print(idx)

# Invalid
idx <- rule_recycle(c(2, 4, 6))
idx
```


