---
title: "Proposed API for Black Box Model Runs"
output: html_document
---

```{r, include=FALSE}
library(dplyr)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "ensemble_model_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
```

# Goals

1. Develop a unified interface for generated PEcAn "forward models".
2. No modifications of any existing low-level PEcAn functionality.

The goal is to provide functionality that works something like this.
```{r, eval=FALSE}
f <- get_forward_model(...)
run_algorithm(f, ...)
```

Here `run_algorithm()` might be from an external package, and doesn't need to
know anything about PEcAn.

## Motivating examples

### Fixed ensemble run

Consider an ensemble run with the following characteristics:  

* The model is run at 3 sites  
* Site-specific initial conditions (IC); 2 member (IC) ensemble  
* Site-invariant parameters; 2 member parameter ensemble

```{r, eval=FALSE}
list(
  site  = list("s1", "s2", "s3"),
  ic    = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)
```

A single ensemble run consists of 12 individual model runs.

```{r, echo=FALSE}
slots <- list(
  site  = list("s1", "s2", "s3"),
  ic    = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)

rule <- get_composite_rule(groups = list(c(1,2), 3),
                           rules = list(rule_recycle, rule_match))
idx <- rule(vapply(slots, length, integer(1)))
instantiate_slot_grid(idx, slots) %>% mutate(across(everything(), as.character))
```

### Ensemble run with free parameters

Consider the same above run template with the following modifications:  

* Want to learn the parameters  
* Want to learn the initial condition at site 2

The run spec would now look something like this: 

```{r, eval=FALSE}
list(
  site  = list("s1", "s2", "s3"),
  ic    = list("FREE", "ic2_1", "ic3_1", "FREE", "ic2_2", "ic3_2"),
  param = list("FREE")
)
```

If we could generate a forward model `f` of the form

```{r, eval=FALSE}
f(free_inputs) = observable
```

the we could utilize `f` is standard MCMC or optimization software, with the
complicated PEcAn ensemble run mechanisms hidden under the hood.


# Benefits
1. Make PEcAn more modular and extensible
2. Abstraction: separation of algorithm code and run model code
3. Easier debugging and unit testing
4. Interface with external packages
5. Less boilerplate
6. Support both PEcAn models and "R models"

# Overview
1. Formalize `ModelInput`
2. Formalize `EnsembleInput` 
3. Define model runner APIs
4. Define a "forward model generator"

# Formalizing `ModelInput`

## Current PEcAn Approach
* Single model input effectively defined by `(settings, runtime_inputs)`  
    - e.g., `runtime_inputs` typically includes parameters and/or initial conditions.  
* `settings` also stores information beyond model inputs (e.g., parallelization settings)  
* More specifically, single model input defined by something like `(settings$run, defaults, trait.values, ic)`
* Questions:  
    - Is the format for `trait.values` standardized? (e.g., via validators or documentation)
    - Same for `defaults`?

## A simple `ModelInput` abstraction

* Simply a light container with "input slots"
* Includes basic validation (e.g, slots must have names)
* Up to a particular model to validate that the required slots are present

```{r}
input <- ModelInput(param    = c(a=1, b=2),
                    ic       = c(ic1=0, ic2=0.5),
                    settings = list(run = list(a="/file/path")),
                    metadata = list(input_id = 1L))

print(input)
```


# Formalizing `EnsembleInput`

Define an *ensemble model run* as a set of runs of a fixed model with 
different inputs.

## Current PEcAn Approach
* Store list of `settings` objects, one per run `settings` (`MultiSettings`)  
* `MultiSettings` then fed to functions like `write.ensemble.configs()` and `runModule.run.write.configs()`.  
* Questions:  
    - Support for free inputs?  
    - How are free inputs lined up with information in `MultiSettings`?  
* Workflow appears to be geared towards fixed and randomized ensemble runs  
    - Config functions noted above perform sub-sampling (randomization)  
    - Not aware of an interface for ensemble runs not involving sub-sampling  

## An `EnsembleInput` abstraction

* Simply an ordered collection of `ModelInput` objects, with metadata (e.g., run IDs)  
* Can be represented in 3 underlying data structures:  
   1. List of `ModelInputs`s (similar to `MultiSettings`)  
   2. Table of shape `(n_runs, n_slots)` (one row per run)  
   3. Set of unique slot values, plus a broadcasting rule  
   
   
### Broadcast representation
   
```{r}
# Define unique values
slots <- list(
  site = list("s1", "s2", "s3"),
  ic = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)

# Create broadcast rule
rule <- get_composite_rule(groups=list(c(1,2), c(3)), # slot groups: (site, ic), (param)
                           rules=list(rule_recycle, rule_match))

# Instantiate ensemble input object
ens_input <- EnsembleInput(rule, slots)

summary(ens_input)
```


### Table representation

The broadcast rule can be realized at any time to obtain the concrete run 
table. However, this will typically not be necessary as run inputs can be 
generated one at a time when needed.


```{r}
tbl <- as_ensemble_input_table(ens_input)
summary(tbl)

print(tbl %>% mutate(across(everything(), as.character)))
```


# Next steps

## Complete forward model work

1. Define APIs for running model (in progress)  
2. Define forward model generators (basic idea sketched out)  

## Complete standalone tools for calibration  

1. Probability distributions (base classes complete; R6)    
2. Modular MCMC package (core functionality almost complete)  
3. Ensemble Kalman Inversion (need to port over from personal code)  
4. Sensitivity analysis (Meng)








