---
title: "ensemble_input"
output: html_document
---

```{r}
library(dplyr)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(code_dir, "types.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "ensemble_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
```


We define a model ensemble as a set of runs of a particular model at different
inputs. The `EnsembleInput` class provides a standardized container to hold
the set of inputs. This class represents an ordered set of `ModelInput` objects,
which can be stored using one of three underlying data structures: a list,
table, or implicitly via a "broadcast rule". This tutorial introduces 
these representations, and provides various examples of working with 
`EnsembleInput` objects. See the `ModelInput` tutorial for more detailed
information on the individual `model_input.Rmd` objects themselves, and
the `run_model_intro.Rmd` tutorial for information on defining models
that run at one or more model inputs.

# List Representation
The simplest data structure used to encode `EnsembleInput` is a list of 
`ModelInput` objects.

```{r}
mi1 <- ModelInput(list(a=1, b=list(c="2", d=1:3), e=4))
mi2 <- ModelInput(list(a=10, b=list(c="20", d=10*(1:3), e=MetadataSlot("some_metadata"))))

ei <- EnsembleInput(list(mi1, mi2))
summary(ei)
print(class(ei))
```
Notice that the input slots and metadata are not restricted to coincide for
all model inputs comprising the ensemble. The printed summary displays the 
union of all input slots across all inputs. The input-by-input input slot and 
metadata names can also be queried as follows.

```{r}
cat("Input slot names:\n")
slot_names(ei, unique_only=FALSE)

cat("\nMetadata names:\n")
metadata_names(ei, unique_only=FALSE)
```

Each model input is associated with a particular "run" of the model, and 
assigned a corresponding unique `run_id`. We could have manually specified
run IDs by adding a names attribute to the list passed into the 
`EnsembleInput` constructor. Since we didn't do this, default run IDs were
generated for us.

```{r}
run_ids(ei)
```

In addition to uniquely identifying each run, the run IDs define the order
of the `EnsembleInput` object. We can view the inputs for each run on a 
slot-by-slot level by converting to a table format.

```{r}
as_ensemble_input_table(ei)
```

By default, the columns other than `run_id` are list columns, which accommodate
arbitrary structured objects. We can fetch the `ModelInput` corresponding to a 
particular `run_id` using `get_run_input()`.

```{r}
run_input <- get_run_input(ei, "run_1")

summary(run_input)
```

Trying to extract a non-existent run ID will result in an error.

```{r}
try(get_run_input(ei, "run_100"))
```

# Broadcast Representation

The list representation is intuitive and reasonable when one has a pre-defined
set of model inputs. In other cases, one may need to construct a large
ensemble of inputs, where certain slot values may be repeated across many
different runs. In this case, list or table representations require storing
duplicate information. An `EnsembleInputBroadcast` class 
provides a more efficient representation. It is defined by specifying 
a unique set of values in each slot, along with a "broadcast rule" that 
describes how to construct runs by combining the inputs. 

We walk through a sequence of increasingly complicated examples. 

## Recycling along an axis

To start, 
consider an example with two slots, one of length 4 and one of length 2.
We seek to define 4 unique runs by repeating the pattern of the shorter slot
so that it matches the length of the longer one.

```{r}
slot_list <- list(
  a = list(1, 2, 3, 4),
  b = list("one", "two")
)

ei <- EnsembleInput(rule_recycle, slot_list)

cat("Class of created object:\n")
print(class(ei))

cat("\nSummary of object:\n")
summary(ei)

cat("\nVisualizing runs (slot indices):\n")
print(ei$idx_mat)

cat("\nVisualizing runs (slot tags):\n")
get_labeled_idx_mat(ei)
```

The main object defining the runs for a `EnsembleBroadcastInput` is the 
`idx_mat` matrix shown above. The `get_labeled_idx_mat()` helper prepends
the slot names to these indices for improved readability. The print method
for an `EnsembleInputBroadcast` prints this latter visualization.
```{r}
print(ei)
```
The table of actual slot values (rather than values) can also be formed as before.
Recall that the type of columns other than `run_id` defaults to lists.

```{r}
as_ensemble_input_table(ei)
```

As with any
`EnsembleInput` object, the runs can be queried individually. While
`EnsembleInputList` always stores the `ModelInput` objects, the broadcast
representation creates these objects on-demand (i.e., lazily).


```{r}
run_input <- get_run_input(ei, "run_2")

summary(run_input)
print_tree(run_input)
```
`ModelInput`s with a tree structure (depth greater than one) can be accommodated
in `EnsembleInputBroadcast` by setting the names in the slot list to be the
note keys.


```{r}
slot_list <- list(
  `a/b/c` = list(1, 2, 3, 4),
  `a/b/d` = list("one", "two"),
  `a/e` = list(100)
)

ei <- EnsembleInput(rule_recycle, slot_list)

print(ei)

cat("\nRun 2:\n")
print_tree(get_run_input(ei, "run_2"))
```

The above example also illustrates an example where the values in multiple
slots are recycled. Unlike base R, an error will be thrown if the length
of any of the shorter axes does not divide the length of the longest axis
evenly.

```{r}
slot_list <- list(
  `a/b/c` = list(1, 2, 3, 4),
  `a/b/d` = list("one", "two"),
  `a/e` = list(100, 200, 300)
)

try(EnsembleInput(rule_recycle, slot_list))
```

## Cartesian Product

Another pre-defined rule allows taking all possible combinations of slot values.

```{r}
slot_list <- list(
  a = list(1, 2, 3),
  b = list("one", "two"),
  c = list(100)
)

ei <- EnsembleInput(rule_cartesian, slot_list)
```
## Matching across equal-length dimensions
The match rule simply lines up slot values for slots that have the same length.

```{r}
slot_list <- list(
  a = list(1, 2, 3),
  b = list("four", "five", "six"),
  c = list(7, 8, 9)
)

ei <- EnsembleInput(rule_match, slot_list)
print(ei)
```

## Combining Rules
The base rules can be combined to produce more complicated combinations.
In this example, slot `b` is recycled to the length of slot `a`. The Cartesian
product between the resulting six runs is then taken with slot `c`. This results
in a total of 12 runs.


```{r}
slot_list <- list(
  a = list(1, 2, 3, 4, 5, 6),
  b = list("seven", "eight", "nine"),
  c = list(10, 11)
)

rule <- get_composite_rule(groups = list(c(1,2), c(3)),
                           rules = list(rule_recycle, rule_match))

ei <- EnsembleInput(rule, slot_list)
print(ei)
```



