---
title: "Generating Forward Models"
output: html_document
---

```{r echo=FALSE, message=FALSE}
library(dplyr)
library(assertthat)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(code_dir, "types.r"))
source(file.path(code_dir, "batched_array.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "ensemble_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
source(file.path(fwd_dir, "run_model_api.r"))
source(file.path(fwd_dir, "forward_model_generators.r"))
```


## Introduction

In this case, each run requires a parameter vector as input, in addition
to some other information. 

```{r}
slot_list <- list(
  a = list(1, 2, 3, 4, 5, 6),
  b = list("seven", "eight", "nine"),
  c = matrix(7:14, nrow=4L) # Four different values of the parameter vector
)

ei <- EnsembleInputBroadcast(slot_list, 
                             list_rule = list(recycle = c("a", "b")))

# Summary of ensemble
summary(ei)

# Input indices for each run
print(ei)

# Looking at inputs for a particular run
input_slots(get_run_input(ei, "run_20"))
```


```{r}
# Define so that the forward model accepts one value per free slot
# Or multiple such values; i.e., not passing in "value sets".
# If passing one value per slot, there is no need to pass a "new_slot_rule"
# since there is only one way the slots can be combined (matching).
# I guess for now keep the current approach of assuming that the current ei
# stores the "batch dims" for these slots, and we back out the individual value
# dimension.
#
# Allow passing values as:
#  (1) list of arrays
#  (2) list of lists of arrays (all sub-lists must be of equal length)
#  (3) matrix (flat representation - one row per value)
#
# With this setup, there is only one way that the free slots can be broadcast -
# taking a product with the fixed slots. So perhaps it makes more sense to 
# specify a "base" ens_input that only has fixed slots. Any new slots will
# be matched against each other, and then Cartesian product'd against the
# existing fixed slots. This is a fairly natural notion of "vectorized".

gen_vectorized_forward_model(f, base_ens_input,
                             new_slot_dims=list(d=4, e=c(3,3)),
                             output_operator)
```


Suppose we want to generate a forward model that uses this ensemble run as 
the "run template"; that is, each execution of the forward model will run
the underlying model 24 times. Suppose also that we want to be able to pass in
the set of parameter vectors on the fly for each run. In other words,
slots `a` and `b` will be fixed, while slot `c` is free. We first define a toy
model.

```{r}
# Just print the input to visualize what is happening.
f <- function(x) {
  x <- input_slots(x)
  print(paste0(x$a, " - ", x$b, " - ", .get_dims_string(x$c)))
}
```


Now we can generate a "forward model", which returns a function that accepts
only the free slots as arguments. A "matrix forward model" always requires 
a matrix as input. The interpretation of this matrix is completely dependent
on the `EnsembleInput` template we defined previously. The forward model
generator by default prints a description of the forward model, which can
be silenced by setting `verbose = FALSE`.

```{r}
fwd <- gen_matrix_fwd_model(f, ei, slot_names="c")
```

Here's what the forward model looks like when printed. It is simply an R 
function that uses a closure, which allows the function to have a state 
determined by the `EnsembleInput` template, free slot names, and some other
information used for validating inputs to the forward model and using them
to set the values of the free slots.

```{r}
print(fwd)
```
Let's start by making sure the forward model behaves as expected by running it
at the existing value of `c` we set above, and comparing the result to a standard
call to `run_model_ensemble()`.


```{r}
output1 <- run_model_ensemble(f, ei)
```


```{r}
output2 <- fwd(slot_list$c)
```

```{r}
# Check that the two ensemble run outputs are the same.
all.equal(output1, output2)
```

The benefit of the forward model is that we now have a black-box function that
can easily be run at new parameter values without having to think about the 
`EnsembleInput`, model run API, etc.

```{r}
# Run forward model at new parameter values
new_c <- matrix(101:108, ncol=2L)
new_output <- fwd(new_c)
```

Trying to pass invalid inputs will result in errors.

```{r}
try(fwd(NULL)) # Non-array input
try(fwd(1:2))  # Incorrect dimension
```


