---
title: "model_input"
output: html_document
---

```{r}
base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "ensemble_model_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
```


# Representing a Single Model Input

The `ModelInput` class encodes the set of inputs required to run a model.
It is a simple container with elements *slots* and *metadata*, both of which
are named lists (or empty lists). The slots correspond to different pieces of 
the input (e.g., parameters, initial conditions, a filepath to data, etc.)

Let's start by considering a model input consisting of two slots: 
(1) `param`, a vector of scalar parameters, and (2) `data_path`, a file path 
to some required input data.

```{r}
input <- ModelInput(param = c(a=1, b=2),
                    data_path = "path/to/data",
                    metadata = list(input_id = 1L))
print(input)
print(class(input))
```

Helper functions and methods are defined to extract slots, metadata, and their 
names.

```{r}
slot_names(input)
metadata_names(input)
slots(input)
metadata(input)
```

Attempting to create unnamed slots (or metadata) will result in an error,

```{r}
input <- try(ModelInput(param=2, "unnamed_slot"))
print(input)
```
but empty inputs are allowed.

```{r}
input <- ModelInput()
print(input)
```

# Representing a Collection of Model Inputs

A collection of model inputs can be combined into an `EnsembleInput` object.
This is the primary building block for constructing vectorized, or *ensemble*
model runs. `EnsembleInput` objects can be stored in three different
formats, which each may be preferred in certain contexts. A common constructor
can be used to generate these objects in the three formats, and methods are 
provided to easily convert between the three. We discuss each representation
below.

## List Representation

The simplest data structure used to encode `EnsembleInput` is a list of `ModelInput` objects.

```{r}
input1 <- ModelInput(param = c(a=1, b=2),
                     data_path = "path/to/data")
input2 <- ModelInput(param = c(a=10, b=20),
                     data_path = "path/to/data2")

ens_input <- EnsembleInputList(list(input1, input2))
print(ens_input)
```


The inputs are not required to all have the same input slots. We can access
the union of the slots across all inputs, or a list of the slots for each input.

```{r}
input1 <- ModelInput(param = c(a=1, b=2),
                     data_path = "path/to/data",
                     metadata=list(extra_info = "info"))

input2 <- ModelInput(param = c(a=10, b=20),
                     data_path = "path/to/data2",
                     ic = list(ic1=0, ic2=100, ic3=50))

ens_input <- EnsembleInput(list(input1, input2))
print(ens_input) # By default, prints the union.

# Print union of slot names
slot_names(ens_input)

# Print list of slot names for each input
slot_names(ens_input, unique_only=FALSE)

# Same deal with metadata
metadata_names(ens_input)
metadata_names(ens_input, unique_only=FALSE)
```

Each input/run is associated with a run ID. We could have specified these 
above by passing in a named list `list(run1=input1, run2=input2)` to the 
`EnsembleInput` constructor. If no names are specified, the constructor 
sets default run IDs.

```{r}
run_ids(ens_input)
```

The `length` method for an `EnsembleInput` is defined so that it returns the
total number of runs.

```{r}
length(ens_input)
```

## Table Representation
```{r}

```





