---
title: "model_input"
output: html_document
---

```{r}
library(dplyr)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "ensemble_model_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
```


# Representing a Single Model Input

The `ModelInput` class encodes the set of inputs required to run a model.
It is a simple container with elements *slots* and *metadata*, both of which
are named lists (or empty lists). The slots correspond to different pieces of 
the input (e.g., parameters, initial conditions, a filepath to data, etc.)

Let's start by considering a model input consisting of two slots: 
(1) `param`, a vector of scalar parameters, and (2) `data_path`, a file path 
to some required input data.

```{r}
input <- ModelInput(param = c(a=1, b=2),
                    data_path = "path/to/data",
                    metadata = list(input_id = 1L))
print(input)
print(class(input))
```

Helper functions and methods are defined to extract slots, metadata, and their 
names.

```{r}
slot_names(input)
metadata_names(input)
slots(input)
metadata(input)
```

Attempting to create unnamed slots (or metadata) will result in an error,

```{r}
input <- try(ModelInput(param=2, "unnamed_slot"))
print(input)
```
but empty inputs are allowed.

```{r}
input <- ModelInput()
print(input)
```

# Representing a Collection of Model Inputs

A collection of model inputs can be combined into an `EnsembleInput` object.
This is the primary building block for constructing vectorized, or *ensemble*
model runs. `EnsembleInput` objects can be stored in three different
formats, which each may be preferred in certain contexts. A common constructor
can be used to generate these objects in the three formats, and methods are 
provided to easily convert between the three. We discuss each representation
below.

## List Representation

The simplest data structure used to encode `EnsembleInput` is a list of `ModelInput` objects.

```{r}
input1 <- ModelInput(param = c(a=1, b=2),
                     data_path = "path/to/data")
input2 <- ModelInput(param = c(a=10, b=20),
                     data_path = "path/to/data2")

ens_input <- EnsembleInput(list(input1, input2))
summary(ens_input)
```

The inputs are not required to all have the same input slots. We can access
the union of the slots across all inputs, or a list of the slots for each input.

```{r}
input1 <- ModelInput(param = c(a=1, b=2),
                     data_path = "path/to/data",
                     metadata=list(extra_info = "info"))

input2 <- ModelInput(param = c(a=10, b=20),
                     data_path = "path/to/data2",
                     ic = list(ic1=0, ic2=100, ic3=50))

ens_input <- EnsembleInput(list(input1, input2))
summary(ens_input)

# Print union of slot names
slot_names(ens_input)

# Print list of slot names for each input
slot_names(ens_input, unique_only=FALSE)

# Same deal with metadata
metadata_names(ens_input)
metadata_names(ens_input, unique_only=FALSE)
```

Each input/run is associated with a run ID. We could have specified these 
above by passing in a named list `list(run1=input1, run2=input2)` to the 
`EnsembleInput` constructor. If no names are specified, the constructor 
sets default run IDs.

```{r}
run_ids(ens_input)
```

The number of runs and number of slots can be queried using `n_runs()`
and `n_slots()`. The `dim` method returns a tuple containing these two numbers.

```{r}
n_runs(ens_input)
n_slots(ens_input)
dim(ens_input)
```

## Table Representation

The ensemble input can alternatively be stored in a table, with one row per 
model input (i.e., one row per run). The column name prefixes `slot_`
and `metadata_` are used to identify slot and metadata columns, respectively.
A special column named `run_id` stores the associated run IDs. All metadata
and slot columns are stored as "tibble" list columns, which allows storing
arbitrary structured objects.


```{r}
ens_input <- as_ensemble_input_table(ens_input)

summary(ens_input)
print(class(ens_input))
```

The same methods for accessing slot names, metadata names, and run IDs work when 
the ensemble input is in the table format.

```{r}
slot_names(ens_input)
metadata_names(ens_input)
run_ids(ens_input)

n_runs(ens_input)
n_slots(ens_input)
dim(ens_input)
```
An `EnsembleInputTable` object is a tibble, so we can dispatch standard tibble
and data.frame methods for manipulating the table.

```{r}
print(ens_input)
```

## Broadcast Representation

The downside of both the table and list representations are they both
store redundant information when some slot values are utilized in multiple runs.
The broadcast representation addresses this shortcoming, storing only a unique
set of values for each slot, along with a broadcast rule that encodes how to 
construct the runs from the slot values.

```{r}
slots <- list(
  site = list("s1", "s2", "s3"),
  ic = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)

# Create new broadcast rule
rule <- get_composite_rule(groups=list(c(1,2), c(3)), # slot groups: (site, ic), (param)
                           rules=list(rule_recycle, rule_match))

ens_input <- EnsembleInput(rule, slots)

summary(ens_input)
print(class(ens_input))
```

The runs are internally defined by an matrix of indices.

```{r}
ens_input$idx_mat
```

This is easier to understand with some text labels attached in place of the 
integer indices.

```{r}
get_labeled_idx_mat(ens_input)
```

We can convert to an `EnsembleInputTable` to fill the table with the actual 
slot values, rather than string labels for each value. 

```{r}
tbl <- as_ensemble_input_table(ens_input)

summary(tbl)
head(tbl)
```







