---
title: "pecan_run_model"
output: html_document
---


```{r echo=FALSE, message=FALSE}
# PEcAn dependencies
library(PEcAn.settings)
library(PEcAn.workflow)
library(PEcAn.logger)
library(PEcAn.utils)
library(PEcAn.remote)
library(PEcAn.uncertainty)
```


```{r echo=FALSE, message=FALSE}
library(dplyr)
library(assertthat)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(code_dir, "types.r"))
source(file.path(code_dir, "batched_array.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "pecan_model_input.r"))
source(file.path(fwd_dir, "run_model_api.r"))
source(file.path(fwd_dir, "run_pecan_model_api.r"))
source(file.path(fwd_dir, "ensemble_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
```


This tutorial provides an introduction to a high-level interface for running
PEcAn models. This interface is built as a wrapper around the standard PEcAn
workflow for running models, and is designed to facilitate (1) easy construction
of complex model ensembles and (2) the ability to construct standard R functions
that enable running PEcAn models within external R packages.

We utilize the [SIPNET](https://github.com/PecanProject/pecan/tree/develop/models/sipnet)
model as an example. To actually run this code, one must save valid 
`.nc` files to the paths specified below. In particular, it is 
assumed that IC files are stored in a sub-directory of `ic_dir` with naming convention
`<site_id>/IC_site_<site_id>_<ensemble_id>.nc`. Likewise, the driver files should be
located in a sub-directory of `met_dir` with naming convention 
`ERA5_<site_id>_<ensemble_id>/ERA5.<ensemble_id>.<year_range>.clim`. In this tutorial, 
we use the year range `2012-01-01.2024-12-31`. 
We use site IDs 1-3 and ensemble IDs 1-2. The helper functions
below are used for constructing filepaths given site and ensemble ID.


```{r}
base_out_dir <- "/projectnb/dietzelab/arober/sipnet_calibration/runs/api_tutorial"
ic_dir <- "/projectnb/dietzelab/dongchen/anchorSites/NA_runs/IC/IC"
met_dir <- "/projectnb/dietzelab/dongchen/anchorSites/NA_runs/ERA5_2012_2024"
```

```{r}
get_ic_path <- function(site_id, ens_id) {
  filename <- paste0("IC_site_", site_id, "_", ens_id, ".nc")
  file.path(ic_dir, site_id, filename)
}

get_met_path <- function(site_id, ens_id) {
  subdir <- paste("ERA5", site_id, ens_id, sep="_")
  filename <- paste("ERA5", ens_id, "2012-01-01.2024-12-31", "clim", sep=".")
  file.path(met_dir, subdir, filename)
}
```

We'll also be needing a bit of site information. To keep things relatively
self-contained, we hard code some information about the sites here.

```{r}
site_coords <- list(
  site1 = list(lat=82.5458343505859, lon=-24.5625010172526),
  site2 = list(lat=82.5375010172526, lon=-22.6625010172526)
)


get_site_info <- function(site_id) {
  site_id <- as.integer(site_id)
  
  list(
    id = site_id,
    lat = site_coords[[site_id]]$lat,
    lon = site_coords[[site_id]]$lon,
    met.start = "2012/01/01",
    met.end =  "2024/12/31"
  )
}
```


   <site>
    <id>8000</id>
    <lat>7.0125010172526</lat>
    <lon>-71.4208343505859</lon>
    <name>weighted_sample</name>
    <met.start>2012/01/01</met.start>
    <met.end>2024/12/31</met.end>
   </site>


# Executing a Single Model Run

We start by walking through how to set up and execute a single forward model
run for a PEcAn model. The traditional PEcAn workflow consists of specifying
most of the input information required by a model run in a 
[Settings](https://github.com/PecanProject/pecan/blob/develop/base/settings/R/Settings.R)
object. In the alternative workflow presented here, we retain the use of this
object but view `Settings` primarily as a container for (1) settings
that will be fixed throughout the entirety of an analysis, and (2) default
model inputs. This includes metadata, model setup, filepaths, and fixed input
data. We introduce a `ModelInput` object which is specifically designed to 
hold model inputs.

## Model Setup

We start by specifying the model we will be using.

```{r}
model_settings <- list(
  id = "1000000022",
  type = "SIPNET",
  revision = "ssr",
  delete.raw = "FALSE",
  binary = "/projectnb/dietzelab/dongchen/SIPNET/sipnet",
  jobtemplate = "~/pecan/models/template/inst/template_geo.job"
)
```


## Fixed model inputs

In this initial simple example, we will fix a single initial condition and 
a single driver trajectory.

```{r}
input_settings <- list(
  met = list(path = get_met_path(1,1)),
  poolinitcond = list(source = "RS_veg",
                      output = "poolinitcond",
                      path = get_ic_path(1,1))
)
```


## `Settings` object

We now wrap the above information in a PEcAn `settings` object.

```{r}
settings_list <- list(
  
  # Directory information
  outdir = base_out_dir,
  modeloutdir = file.path(base_out_dir, "out"),
  rundir = file.path(base_out_dir, "run"),
  
  # Run model locally
  host = list(name="localhost"),
  
  # PFTs information
  pft = list(
    plant = list(name="temperate.deciduous.HPDA", constants=list(num="1")),
    soil = list(name="soil.HPDA", constants=list(num="1"))
  ),
  
  # SIPNET setup
  model = model_settings,
  
  # Model inputs, site, and time horizon
  run = list(
    site = get_site_info(1),
    inputs = input_settings,
    start.date = "2012/01/01",
    end.date = "2014/12/31"
  )
  
)


settings <- PEcAn.settings::as.Settings(settings_list)
```

## Overwriting Default Inputs

We now introduce the `ModelInput` class, which is simply a nested list/tree-like
data structure acting as a container for model inputs and additional metadata.
This is similar in structure to the `settings` object, but adds additional
structure and functionality. Each leaf of a `ModelInput` tree is either 
considered a model input or metadata. For example, we can wrap the settings
list from above as a `ModelInput`.

```{r}
print_tree(ModelInput(settings_list))
```

As seen above, by default the `ModelInput` constructor treats all leaves
as inputs, unless they are explicitly tagged as metadata. `ModelInput` knows
nothing about PEcAn; it is simply an implementation of a tree data
structure geared towards representing model inputs. `ModelInput` interfaces
with the existing PEcAn workflow by adopting the following philosophy:

1. `settings` should contained fixed inputs/metadata, as well as default values
that may be overwritten on any particular model run.  
2. `ModelInput` should specifically contain model inputs which will be used
both to overwrite default values in `settings`. This includes both directly 
overwriting `settings` values, as well as passing arguments (e.g., trait values)
to the model's write config function.

As initial example, let's suppose we want to execute a model run using the 
above defaults in `settings`, but change two things: (1) switch the ensemble
member for the met data, and (2) manually specify the values of 
certain parameters. Addressing (1) requiring overwriting the met
path in `settings`, while addressing (2) requires passing the parameter
values as an argument to SIPNET's `write.config` function. Let's set up a 
`ModelInput` object that contains these two modifications of the model inputs.

```{r}
# Values to override in settings object
settings_override <- .assign_value_at_path(list(), "run/inputs/met/path", 
                                           get_met_path(1,2))

# Values to pass to write config function
trait_values <- InputSlot(list(c(root_turnover_rate=0.5, Amax=100)))
param_override <- list(trait.values = trait_values)

# PecanModelInput stores both of these overrides as separate sub-trees
model_input <- PecanModelInput(config_tree = param_override,
                               settings_tree = settings_override)

print_tree(model_input)
```

We require a way to differentiate which values should override PEcAn settings
values and which will be passed to the write config function. As seen above,
the PEcAn model input object addresses this by storing two specially designated
sub-trees for these purposes.

We can investigate the values that will be passed to the write config function.
```{r}
# These names should match config argument names
config_arg_names(model_input)

# These are the values that will passed to each argument
config_args(model_input)
```

It looks a bit awkward that `trait.values` is wrapped in a list, but this 
is the required format for the `trait.values` argument in `write.config.SIPNET()`.
The list returned by `config_args(model_input)` will be passed directly to this
function, so the values must be formatted as required by the write config.
Note that when we created the `model_input` tree we also wrapped the 
`trait.values` list in `InputSlot()`. This is due to the fact that by default
the tree interprets lists as branches, not leaves. To force an object to 
be interpreted as a leaf, we can wrap it in either `InputSlot()` or
`MetadataSlot()`. The latter will never be used when overwriting default values,
but may be useful for storing auxiliary information.

Next, we investigate the values that will be used to overwrite values in the 
PEcAn settings object.

```{r}
# These key paths will be altered in the settings object
settings_keys(model_input)

# These are the values that will be set at the key paths
settings_input_slots(model_input)
```
In our current example only one value will be overwritten. As we will see, 
when running the model the input slot overwriting is performed behind the 
scenes. We can also perform this manually if desired. 

```{r}
# Update values in settings that are found in the settings sub-tree of model_input 
new_settings <- update_pecan_settings(settings, model_input)

# The met path has been updated.
new_settings$run$inputs$met$path
```

## Running the model

The `run_model()` S3 generic provides a unified interface for running different
kinds of models. If called like `run_model(settings, model_input, ...)` where
the first argument is a PEcAn `settings` object then the `run_model.Settings()`
method will be dispatched, which acts as a wrapper around the standard PEcAn
model run workflow. When this is called, `model_input` will be used both to
override defaults in `settings` and to pass arguments to the write config function.

```{r}
output <- run_model(settings, model_input, run_id="test_run", 
                    variables="GPP", print_summary=FALSE)

print(head(drop(output$GPP)))
```

Additional named arguments (e.g., `variables` and `print_summary` above)
are forwarded to `PEcAn.utils::read.output()`, which handles how the model
output is read from file. If one wishes to separate the process of running 
a model into distinct steps of writing configs, executing the model, and
then reading the output the functions `prep_pecan_model_run()` and
`start_pecan_model_run()` may be used, which is what `run_model()`
does under the hood.

# Ensemble Runs








