---
title: "run_model_intro"
output: html_document
---

```{r}
library(dplyr)

base_dir <- "/projectnb/dietzelab/arober/pecan/modules/prob.tools"
code_dir <- file.path(base_dir, "R")
fwd_dir <- file.path(code_dir, "forward_model")

source(file.path(code_dir, "utils.r"))
source(file.path(code_dir, "types.r"))
source(file.path(fwd_dir, "broadcast_rules.r"))
source(file.path(fwd_dir, "model_input.r"))
source(file.path(fwd_dir, "run_model_api.r"))
source(file.path(fwd_dir, "ensemble_input.r"))
source(file.path(fwd_dir, "ensemble_input_list.r"))
source(file.path(fwd_dir, "ensemble_input_table.r"))
source(file.path(fwd_dir, "ensemble_input_broadcast.r"))
```

This vignette provides an introduction to the standardized interface for
running models. A *model* is viewed generically as an operation that maps
inputs to outputs. This tutorial describes:
1. The `ModelInput` class  
2. The `run_model()` interface  
3. The use of `EnsembleInput` and `run_model_ensemble()` to vectorize a model
over multiple inputs.

# A Standard Representation of Model Inputs

A model input is generically viewed as a container, possessing multiple 
input types; i.e., "slots". This broad definition is encoded by the 
`ModelInput` class.

We start by considering a model input consisting of two slots, which store 
a numeric scalar and square matrix, respectively. The model input object
can also store associated metadata via a named list.

```{r}
metadata_list <- list(note = "some metadata")

model_input <- ModelInput(x = 1.0, 
                          y = diag(rep(1,2)),
                          metadata=metadata_list)

print(model_input)
```

The input values for the various input slots can be accessed via `slots()`,
which returns a list of the slots.

```{r}
slots(model_input)
```

The metadata is similarly accessed via `metadata()`.

```{r}
metadata(model_input)
```

These values can also be obtained using the list notation `model_input$slots`
and `model_input$metadata`.
The `slot_names()`, `metadata_names()`, and `n_slots()` methods also query 
information about the model input.

```{r}
slot_names(model_input)
metadata_names(model_input)
n_slots(model_input)
```

A new slot may be added after a model input is already created via the
`add_slot()` method. The new slot name must not conflict with an existing
slot name.

```{r}
try(add_slot(model_input, "x"))
```

Suppose we want to extend our model to have a third input `z`. We can add this 
slot, and immediately set its value as follows.

```{r}
model_input <- add_slot(model_input, "z", 2.0)

print(model_input)
print(paste0("Value of new slot: ", slots(model_input)$z))
```

# Running a Model
A model is simply defined as an operation that takes in a `ModelInput`
object and returns an output. The `run_model(model_obj, model_input, ...)`
generic provides a standardized API, for which methods can be defined
for particular run model mechanisms. In this tutorial, we discuss the 
simplest setting in which `model_obj` is an R function. Custom methods 
can be defined for more complicated cases where the model execution requires
running external code, interacting with a computing cluster, etc.

The `run_model.function` method is already defined, so we need only define
the R function that represents the model map. We consider a toy example here.

```{r}
f <- function(model_input) {
  val <- slots(model_input)
  val$x + val$y + val$z
}
```

We can now run the model at the input we defined earlier.

```{r}
run_model(f, model_input)
```


# Model Ensemble
We define a model ensemble as a set of runs of a particular model at different
inputs. The `EnsembleInput` class provides a standardized container to hold
the set of inputs. This class represents an ordered set of `ModelInput` objects,
which can be stored using one of three underlying data structures: a list,
table, or implicitly via a "broadcast rule". We briefly introduce these
representations here. The functionality built into `EnsembleInput` is 
explained in more depth in other tutorials.

## List of Model Inputs

The simplest data structure used to encode `EnsembleInput` is a list of `ModelInput` objects.

```{r}
model_input_2 <- ModelInput(x = 2.0, 
                            y = 2*diag(rep(2,2)),
                            z = 4.0)

ens_input <- EnsembleInput(list(model_input, model_input_2))
summary(ens_input)
print(class(ens_input))
```
Each model input is associated with a particular "run" of the model, and 
assigned a corresponding unique `run_id`. We could have manually specified
run IDs by adding a names attribute to the list passed into the 
`EnsembleInput` constructor. Since we didn't do this, default run IDs were
generated for us.

```{r}
run_ids(ens_input)
```

In addition to uniquely identifying each run, the run IDs define the order
of the `EnsembleInput` object. The default method for generating 
ensemble runs will perform each run serially in this order.

```{r}
run_model_ensemble(f, ens_input)
```
We can view the inputs for each run on a slot-by-slot level by converting
to a table format.

```{r}
as_ensemble_input_table(ens_input)
```

By default, the columns other than `run_id` are list columns, which accomodate
arbitrary structured objects. 

The `ModelInput` objects within an `EnsembleInput` are not required to all 
have the same slots. Let's consider adding another input that lacks the
`z` slot. We will have to also update our previous model, which assumed
the `z` slot was present. We also add custom run IDs this time.

```{r}
model_input_3 <- ModelInput(x = 0.0, 
                            y = diag(rep(2,2)))

f <- function(model_input) {
  val <- slots(model_input)
  Reduce(sum, val)
}


ens_input <- EnsembleInput(list(id1 = model_input, 
                                id2 = model_input_2, 
                                id3 = model_input_3))
summary(ens_input)
print(run_ids(ens_input))
```

The summary is telling us that there are 3 total slots, and the printed slots
represent the union of all slots over the inputs. We can view input-by-input
slot information as follows.

```{r}
slot_names(ens_input, unique_only=FALSE)
```
We can fetch the `ModelInput` corresponding to a particular `run_id` using 
`get_run_input()`.

```{r}
run_id3 <- get_run_input(ens_input, "run_1")

print(run_id3)
print(run_id3$slots)
```

Here are the results of the new model ensemble run.

```{r}
run_model_ensemble(f, ens_input)
```


## Broadcasted Model Inputs

The list representation is intuitive and reasonable when one has a pre-defined
set of model inputs. In other cases, one may need to construct a large
ensemble of inputs, where certain slot values may be repeated across many
different runs. In this case, list or table representations are inefficient
as they store this duplicate information. An `EnsembleInputBroadcast` class 
provides a more efficient representation. It is defined by specifying 
a unique set of values in each slot, along with a "broadcast rule" that 
describes how to construct runs by combining the inputs. 

Let's consider an example where a model is run at different spatial 
locations (sites), and the model requires a site-specific input. Let's
also suppose that we want to run ensembles at each site in which model
parameters and initial conditions are varied. The parameter ensemble is
site-invariant, while the initial condition ensemble varies by site. 
In place of real inputs, we'll store all of this information as strings
for easy visualization later. We use the convention `ici_j` for the 
`jth` member of the initial condition ensemble at site `i`.

```{r}
slot_list <- list(
  site = list("s1", "s2", "s3"),
  ic = list("ic1_1", "ic2_1", "ic3_1", "ic1_2", "ic2_2", "ic3_2"),
  param = list("p1", "p2")
)

# Create new broadcast rule
rule <- get_composite_rule(groups=list(c(1,2), c(3)), # slot groups: (site, ic), (param)
                           rules=list(rule_recycle, rule_match))

ens_input <- EnsembleInput(rule, slot_list)

summary(ens_input)
print(class(ens_input))
```
The main object stored by an `EnsembleInputBroadcast` is an index matrix
that assigns slot values to particular runs.

```{r}
ens_input$idx_mat
```

Attaching string labels makes this easier to understand. The print method
of `EnsembleInputBroadcast` provides exactly this.

```{r}
ens_input
```

We can run a model ensemble exactly as before. Let's define a forward model that 
simply appends the input strings, so that we can easily visualize the output.


```{r}

f <- function(model_input) {
  slot_list <- slots(model_input)
  paste0(unlist(slot_list), collapse=" | ")
}

run_model_ensemble(f, ens_input)
```




